Imminent range shifts in transboundary fish stocks challenge their management under climate change

### 2021-03-29 Scripts/Manuscript.Rmd
Included first round of co-authors final comments

Most fisheries management is poorly adapted to address species’ shifting distributions as a result of human driven warmer, more acidic and less oxygenated ocean waters. Over the past century, human activities have dramatically altered the physical and biogeochemical conditions of the ocean, resulting in warmer, more acidic and less oxygenated waters [@IPCC:2019tn]. Marine species’ distributions reflect species’ preferences for discrete environmental conditions [@Hutchinson:1957nt]. As a result of changing ocean conditions, many marine species have shifted their distributions towards higher latitudes, deeper water or following local temperature gradients to remain within their optimal environmental “niche” [@Poloczanska:2016kk]. The biogeography of marine species is projected to continue to shift as ocean conditions change through the 21st century [@Cheung:2010dt], impacting fisheries production and economies [@Sumaila:2019ew], and compromising our capacity to reach international sustainability goals within the 2030 Agenda for sustainable development such as Goal 14 - life below water [@UnitedNations2018; @Singh:2017ds; @Pecl:2017hu]. The projected risks of impacts can be reduced by improving the effectiveness of current fisheries management, including for species that cross international borders, also known as ‘shared stocks’ [@Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by Coastal States [@UnitedNations:1986tl]. As defined by the United Nations' Food and Agriculture Organization, shared stocks can be classified into four non-exclusive categories: (*i*) transboundary stocks, which cross neighboring EEZs; (*ii*) straddling stocks that, in addition to neighboring EEZs, also visit the adjacent high seas; (*iii*) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and (*iv*) discrete stocks that are only present in the high seas [@Munro:2004th]. This study focuses on transboundary stocks exploited by fisheries operating within EEZs. Countries are responsible for the management of stocks within their EEZs. Under UNCLOS, EEZs are encouraged to cooperate when stocks are shared [@UnitedNations:1986tl]. A recent study estimates that there are 633 transboundary fish species globally, representing 67% of identified exploited taxa. Between 2005 and 2010, these species yielded an annual average of 48 million tonnes of catch and USD 78 billion in fishing revenue [@PalaciosAbrantes:2020tv].

The effectiveness of fisheries management for transboundary species is challenged by shifts in stocks’ distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by; @Pinsky:2020jx; @Oremus:2020id]. In most cases, catch or fishing effort quotas for transboundary stocks are based on historical records and do not necessarily consider the full distribution range of the stock [@FredstonHermann:2018kp], nor the effects of climate change on fish stocks and associated fisheries [@PalaciosAbrantes:2020tv; @Sumby:2021gw]. While many transboundary stocks experience natural seasonal spatial variation (e.g., migration, reproductive cycle), misalignment between fisheries resources allocation and distributional shifts beyond such natural variation have previously resulted in unsustainable fishing levels and international disputes [@Miller:2013iv; @Spijkers:2017ij; @Crespo:2020cs]. Such patterns are expected to be exacerbated under continuing climate change [@Pinsky:2018cb; @Sumaila:2011bq]. The importance of transboundary stocks for both catch and fisheries revenue [@PalaciosAbrantes:2020tv] highlights their critical role in achieving the  Sustainable Development Goals (SDG) under the United Nations 2030 Agenda. Understanding when climate change will affect the sharing dynamics of transboundary stocks and the intensity of the resulting impacts is important for developing climate resilient international ocean governance and achieving the goals set out under the 2030 Agenda [@Pinsky:2018cb; @Link:2010ei; @Sumaila:2020tp].

Here, we employ a mechanistic population dynamic model driven by output from a comprehensive Earth System Model with ten ensemble members to project transboundary stocks’ distribution across the 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario (RCP8.5; Methods). We then estimate the year in which the 663 transboundary species (9,132 stocks) are projected to shift their shared distribution beyond historical variation and the intensity of this shift (i.e., time of emergence; see Methods). We treat each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population (15) and only consider shared stocks between neighboring EEZs (i.e., excluding the high seas).

@@

Our results suggest that 4,119 transboundary stocks will experience a range shift beyond historical variability as early as 2006 and up to 2100 (Fig. \@ref(fig:toeFigure1)A). This corresponds to 68% of the studied stocks (Fig. \@ref(fig:toeFigure1)C).The average year of shift in the shared distribution of these stocks in all EEZs analyzed is projected to be 2036 (s.d. of all species, $\pm$ 28 years). Furthermore, 96% of the world’s EEZs will experience at least one transboundary stock shift between 2006 and 2100 with around 80% having recorded a distributional shift in stocks by the UN 2030 Agenda’s deadline (Fig. \@ref(fig:toeFigure1)C).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "The timeline is presented according to cumulative A. Number of stocks (cumulative), B. Number of EEZs and C. proportion of total stocks and EEZs. Dashed lines represent the year by which countries have committed to reach full implementation of the 2030 Agenda, as well as mid and late change  in range shift (Methods), respectively."}

@@

<!-- ![Figure 1. Cumulative increase in time of emergence by Exclusive Economic Zone and transboundary species in absolute numbers and percentage.](../Figures/Fig2.png) -->

The median year in which transboundary stocks experience a range shift varied significantly according to the geographic region of the neighboring EEZs (Methods; Kruskal-Wallis, $X^2$ = 242.11, DF = 93, *p* < 0.001; Figure \@ref(fig:toeFigure2)A)) and a species’ habitat association (Methods; *p* < 0.05; Table S\@ref(tab:toeTableS2)). Overall, most tropical EEZs and stocks will see earlier shifts, with the EEZs of Latin America, the Caribbean and Polynesia experiencing shifts significantly earlier (*p* < 0.05; see Tables S\@ref(tab:toeTableS1, S\@ref(tab:toeTableS2 for full statistics) than almost any other region (Fig. \@ref(fig:toeFigure2)B, \@ref(fig:toeFigureSX)). In contrast, EEZs and stocks located in temperate regions like northern Europe and eastern Asia are projected to experience much later range shifts (Table \@ref(tab:toeTableS2)).Stocks in countries such as Brazil and New Zealand are not projected to experience range shifts beyond natural variability (Fig. \@ref(fig:toeFigure2)A). On average, shifting transboundary stocks represented 23% (s.d. $\pm$ 22%) of yearly fishing revenue from fisheries targeting such stocks within global EEZs between 2004 and 2010 (Fig. \@ref(fig:toeFigure2)A). However, large variation exists, with shifting stocks representing less than 1% of fishing revenue in some countries (e.g., Ireland) and over 90% in others (e.g., Marshall Islands). Moreover, in some EEZs, while few transboundary stocks are projected to shift beyond historical variability, they still represent a large proportion of revenue derived from fisheries within that EEZ (e.g., Peru).

```{r toeFigure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Year of change in the shared distribution of 4,119 transboundary stocks. A) Land polygon shows the contribution of shifting stocks to the country’s total fishing revenue from transboundary stocks. Exclusive Economic Zone polygons represent average range shifts within them. EEZs with no distributional shifts between 2006 and 2100 are represented in aqua color. B) Year of shared distribution shifts by region according to United Nations sub-regions. Points represent the median year of range shift. Horizontal dashed line represents the year 2030. N = North, S = South, W = West and E = East. Ltn. Ame. and the Car. = Latin America and the Caribbean. Aus and New Z. = Australia and New Zealand"}

@@

We estimated the current (2005 and 2010) contribution of shifting stocks to yearly fishing revenue generated from fisheries targeting transboundary stocks within global EEZs. Within five years of the UN 2030 Agenda, 979 transboundary stocks responsible for 5.3 billion USD in fishing revenue are projected to shift within the EEZs of 76 countries (Fig. \@ref(fig:toeFigure3)); with 4.2 billion USDs accounted by 321 stocks shifting in just eleven countries (Fig. \@ref(fig:toeFigure3)). Most of the remaining countries in the top $75^{th}$ revenue percentile will see early shifts (e.g., before 2050) with the exception of Japan, Korea, Peru and Russia, who are projected to have later shifts (e.g., after 2050). In terms of the number of transboundary stocks shifting from historical variability, Spain (n = 122), France (n = 85), China and Portugal (both n = 73), and Senegal (n = 66), emerge as the top 5 countries (Fig. \@ref(fig:toeFigure3)).

```{r toeFigure3, eval = T,  echo = F, fig.width=10, fig.height=8,fig.pos= "H", fig.cap =  "Average range shift of transboundary stocks per country. Countries are coded according to different colored symbols representative of 5 distinct regions. Horizontal line separating the top 75th percentile of annual fishing revenue. Showing country names for those within the top 75th percentile projected to  have an average shift within 5 years of the 2030 Agenda."}

@@

We based the transboundary index on a stock’s centroid defined as the average grid cells that contain the majority of a stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that our method is robust to the abundance level (e.g., the top 95%) of the stock (Fig. \@ref(fig:toeFigureS3)). In addition, we set an arbitrary threshold of 2 s.d. (95% confidence) to determine the time of emergence of the transboundary index. Sensitivity tests to a lower threshold of 1 s.d. (for a 68% confidence) resulted in a global average shift in the shared distribution of transboundary stocks by 2029 $\pm$ 27, a difference of about 7 years (Fig. \@ref(fig:toeFigureS4)). In total, 5,745 transboundary stocks would shift their shared distributions between 2006 and 2100 considering a less conservative metric, that is, 33% (n = 1626) more shifting stocks.

@@

We identified transboundary stocks that are projected to have changes in stock share ratio beyond an EEZ’s threat point by the early (2020-2040) and mid $21^{st}$ century (2040-2060), relative to the recent past (1951-2005). Findings show that by the early $21^{st}$ century, the global average catch proportion of transboundary stocks that are projected to change their stock share ratio beyond the EEZ’s threat point is 59% $\pm$ 17% (Fig. \@ref(fig:toeFigure5)A). By the early $21^{st}$, 85% (n = 239) of the world EEZs would have experienced such changes. Changes in stock share ratio are expected to increase slightly towards the mid 21st century at both the stock and EEZ level (Fig. \@ref(fig:toeFigure5)A and \@ref(fig:toeFigureS5)).

```{r toeFigure5, eval = T,  echo = F, fig.width=12, fig.height=10,fig.pos= "H", fig.cap =  "Changes in stock share ratio below each country’s threat point by 2030 (2021-2040) relative to 1951-2005. Lines represent the average change in transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those gaining shares (arrowhead). Land polygons represent the percentage of stocks that are projected to change their stock share ratio beyond the identified threat point with higher gains identified in warmer colors (green to yellow). Panel B zooms into specific areas shown within the grey dashed line boundaries in A."}

@@

<!-- ![Figure 1. Changes in stock share ratio below each country’s threat point by 2030 (2021-2040) relative to 1951-2005. Lines represent the average change in transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those gaining shares (arrowhead). Land polygons represent the percentage of stocks that are projected to change their stock share ratio beyond the identified threat point with higher gains identified in warmer colors (green to yellow). Panel B zooms into specific areas shown within the grey dashed line boundaries in A.](../Figures/Fig1.png) -->

The direction and intensity of shafts in transboundary stocks are largely related to regional changes in biogeography and the geometry of the EEZs (Fig. \@ref(fig:toeFigure5)A). In some cases, like the Atlantic and Pacific coasts of Northern and Southern America and the Atlantic coast of Southern Africa, the stock share ratio is expected to follow a projected poleward shift. However, shifts along the coasts of Pacific Central America and West Africa  occur along an equatorial direction. Here, transboundary species follow local oceanographic processes like upwelling zones (e.g., Humboldt current) and patterns of oxygen availability, in addition to temperature gradients. Regions where EEZs are relatively small and have multiple borders (e.g., Caribbean or Northwestern Europe) are expected to have a particularly complex exchange of stock share ratio with no established pattern (Fig. \@ref(fig:toeFigure5)B).

@@

Early shifts in the shared distribution of transboundary stocks concur with previous studies that have detected and attributed changes in marine catch composition [e.g., @Cheung:2013gk; @Frainer:2017ji; @Last:2011fd]. For example, in the early 2000s, Humboldt squid (*Dosidicus gigas*) substantially expanded its geographic range poleward, reaching the coast of Washington state (US) as a response to climatic, oceanographic and ecological changes [@Zeidberg:2007gs]. A new fishery targeting Humboldt squid quickly developed on the heels of the species’ range expansion [@Pinsky:2014by]. In the northeast Atlantic, the Atlantic mackerel (*Scomber scombrus*) fisheries are multi-laterally managed by the EU, Norway, Iceland, Russia, and Denmark (on behalf of the Faroe Islands and Greenland) through the North-East Atlantic Fisheries Commission (NEAFC). However, a range expansion of Atlantic mackerel into Icelandic waters in 2007 resulted in Iceland capturing 6% of the fishery’s total allowable catch and a further 18% in 2008, without consultation with NEAFC, threatening the sustainability of the stock [@Spijkers:2017ij]. These changes resulted in disputes between Iceland and the Faroe Islands, as well as among NEAFC member states [@Spijkers:2017ij]. As transboundary stocks continue to shift, international disputes like that of Atlantic mackerel are likely to increase in frequency requiring a revision of management plans [@Pinsky:2018cb; @Engler:2020bg].

@@

Our results underscore regional "hotspots" of climate risk for transboundary fisheries management that will require the prompt adaptation of management plans (Figures \@ref(fig:toeFigure2)A and \@ref(fig:toeFigure5)A). Such regions, like the Caribbean, are characterized by high levels of warming relative to natural internal variability [@IPCC:2019tn; @Hawkins:2012hc] and high vulnerability of species to warming waters [@IPCC:2019tn]. Moreover, they represent relatively small EEZs that border multiple countries (e.g., the Caribbean). Game theory predicts that the greater the number of negotiating parties, the harder it is for parties to reach an agreement [@Gronbaek:2020ei], thus, making it particularly challenging for countries in these regions to coordinate the management of shifting transboundary stocks.

Management plans that are not prepared to respond to range and abundance shifts will be less resilient to climate change [@Miller:2013iv; @Sumaila:2020tp]. Thus, focusing on these “hotspots'' is key to anticipate a potential increase in fisheries conflicts in the coming years. Identified strategies to cope with changes in the shared proportion of transboundary stocks include strengthening of current cooperative mechanisms and the consideration of side payments [@Tunca:2019if], increased international cooperation [@Miller:2013iv] and management rules that capture distributional shifts [@Pinsky:2018cb]. For example, countries that are Party to the Nauru Agreement (PNA) manage their fisheries under a  Vessel Day Scheme. Because this scheme takes into account shifts in stocks and catches, it is a good example of a management system that is somewhat anticipatory [@Aqorau:2018bh]. However, even strategies that take into consideration  species’ shift within the PNA area will likely also have to deal with the ‘newcomer issue’ that NEAFC is currently facing regarding Atlantic mackerel [@Spijkers:2017ij; @Pinsky:2018cb]. Quota allocation methods based on a stock’s current distribution and/or on fixed-historical proportion need to evolve to be more agile [@Sumaila:2020tp; @PalaciosAbrantes:2020vt]. Fisheries based on historic participation could move towards a dynamic method or a combination of both  [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. However, the transition from historic to dynamic allocations can find strong resistance from stakeholders “losing” benefits from the fishery. Side payments are mechanisms that can provide a solution to the inequality produced by shifts in a stock’s distribution [@Tunca:2019if]. These can, but do not have to, be monetary (e.g., the stock-“winning” state compensates the stock-“losing” state for the proportional shift, a conservation fund paid mainly by one Party) [@Miller:2013iv].

@@

The early projected range shifts of a large number of transboundary stocks can be partially attributed to the parallel global emergence of several ocean variables from historical variability (e.g., SST which is projected to increase historical variability by 2030 in 70% of the global ocean under RCP 8.5 [@Rodgers:2015jq; @Frolicher:2016dg; @Mahlstein:2011fy]. Results from multiple marine ecosystem models show that SST and NPP are the main drivers of changes in species distributions across oceanic basins [@BryndumBuchholz:2019kh; @Lotze:2019ce]. Specifically for the DBEM, SST is the main environmental driver of biomass changes in polar, tropical and upwelling ecosystems while NPP drives temperate regions’ results. The combination of the SST emergence pattern and model characteristics could be partially responsible for the early distributional shift of transboundary stocks in the tropics where marine species live close to their thermal tolerance, making them highly vulnerable to warming waters [@IPCC:2019tn], while also explaining the later, and sometimes non-existent, shift at higher latitudes (Fig. \@ref(fig:toeFigure2)B).

>>>>

Most fisheries management is poorly adapted to address species’ shifting distributions as a result of changing ocean conditions. Over the past century, human activities have dramatically altered the physical and biogeochemical conditions of the ocean, resulting in warmer, more acidic and less oxygenated waters [@IPCC:2019tn]. Marine species’ distributions reflect species’ preferences for discrete environmental conditions [@Hutchinson:1957nt]. As a result of changing ocean conditions, many marine species have shifted their distributions towards higher latitudes, deeper water or following local temperature gradients to remain within their optimal environmental “niche” [@Poloczanska:2016kk]. The biogeography of marine species is projected to continue to shift as ocean conditions change through the 21st century [@Cheung:2010dt], impacting fish stocks, fisheries production and economies [@Sumaila:2019ew]. These impacts are compromising our capacity to reach international sustainability goals under the 2030 Agenda for Sustainable Development such as Goal 14 - life below water [@UnitedNations2018; @Singh:2017ds; @Pecl:2017hu]. The projected risks of impacts can be reduced by improving the effectiveness of current fisheries management, including for species that cross international borders, also known as ‘shared stocks’ [@Pinsky:2018cb].

The concept of shared stocks was developed following the ratification of the United Nations Convention on the Law of the Sea (UNCLOS) and the claiming of Exclusive Economic Zones (EEZs) by Coastal States [@UnitedNations:1986tl]. As defined by the United Nations’ Food and Agriculture Organization, shared stocks can be classified into four non-exclusive categories: (*i*) transboundary stocks, which cross neighboring EEZs; (*ii*) straddling stocks that, in addition to neighboring EEZs, also visit the adjacent high seas; (*iii*) highly migratory stocks, mainly tunas and bill-fishes, that migrate across vast oceanic regions including both the high seas and EEZs; and (*iv*) discrete stocks that are only present in the high seas [@Munro:2004th]. This study focuses on transboundary stocks exploited by fisheries operating within EEZs. Under UNCLOS, countries are responsible for the management of stocks within their EEZs and encouraged to cooperate when stocks are shared  [@UnitedNations:1986tl]. A recent study estimates that there are 633 transboundary fish species globally, representing 67% of identified exploited taxa. Between 2005 and 2010, these species yielded an annual average of 48 million tonnes of catch and USD 78 billion in fishing revenue [@PalaciosAbrantes:2020tv].

The effectiveness of fisheries management for transboundary species is challenged by shifts in stocks’ distribution under climate change [@Pinsky:2018cb; @Pinsky:2014by; @Pinsky:2020jx; @Oremus:2020id].In most cases, catch or fishing effort quotas for transboundary stocks are based on historical records and do not necessarily consider the full distribution range of the stock [@FredstonHermann:2018kp], nor the effects of climate change on fish stocks and associated fisheries [@PalaciosAbrantes:2020tv; @Sumby:2021gw]. While many transboundary stocks experience natural seasonal spatial variation (e.g., migration, reproductive cycle), misalignment between fisheries resources allocation and distributional shifts beyond such natural variation have previously resulted in unsustainable fishing levels and international disputes [@Miller:2013iv; @Spijkers:2017ij; @Crespo:2020cs]. Such patterns are expected to be exacerbated under continuing climate change [@Pinsky:2018cb; @Sumaila:2011bq]. The importance of transboundary stocks for both catch and fisheries revenue [@PalaciosAbrantes:2020tv] highlights their critical role in achieving the Sustainable Development Goals (SDG) under the United Nations 2030 Agenda. Understanding when climate change will affect the sharing dynamics of transboundary stocks and the intensity of the resulting impacts is important for developing climate resilient international ocean governance and achieving the goals set out under the 2030 Agenda [@Pinsky:2018cb; @Link:2010ei; @Sumaila:2020tp].

Here, we employ a mechanistic population dynamic model driven by output from a comprehensive Earth System Model with ten ensemble members to project transboundary stocks’ distribution across the 280 EEZs of 198 coastal countries under a high greenhouse gas emissions scenario (RCP8.5; Methods). We then estimate the year in which the 663 transboundary species (9,132 stocks) are projected to shift their shared distribution beyond historical variation and the intensity of this shift (i.e., time of emergence and stock share ratio; see Methods). We treat each species in an EEZ as a single stock due to the lack of more spatially resolved data to delineate the boundary of a population [@PalaciosAbrantes:2020tv] and only consider shared stocks between neighboring EEZs (i.e., excluding the high seas).

@@

Our results suggest that 4,119 transboundary stocks will experience a range shift beyond historical variability as early as 2006 and up to 2100 (Fig. \@ref(fig:toeFigure1)A). This corresponds to 68% of the studied stocks (Fig. \@ref(fig:toeFigure1)C). The average year of shift in the shared distribution of these stocks in all EEZs analyzed is projected to be 2036 (s.d. of all species, $\pm$ 28 years). Furthermore, 96% of the world’s EEZs will experience at least one transboundary stock shift between 2006 and 2100 with around 80% having recorded a distributional shift in stocks by the UN 2030 Agenda’s deadline (i.e., year 2030) (Fig. \@ref(fig:toeFigure1)C).

```{r toeFigure1, eval = T,  echo = F, fig.width = 9, fig.height=4, fig.pos= "H", fig.cap =  "Timeline of shifts in transboundary shared stock distribution. (A) Number of stocks (cumulative) with their distribution centroids that have shifted beyond natural variations, (B) Number of EEZs (cumulative) with the distribution centroid of at least one stock that have shifted beyond natural variations and (C) proportion of total shifting stocks and EEZs with at least one shifting stock, beyond natural variability. Dashed lines represent the year by which countries have committed to reach full implementation of the 2030 Agenda, as well as mid and late change  in range shift (Methods), respectively."}

@@

The median year in which transboundary stocks experience a range shift varied significantly according to the species’ habitat association (Methods; Kruskal-Wallis, $X^2$ = 203.85, DF = 93, *p* < 0.001; Tables S\@ref(tab:toeTableS1), S\@ref(tab:toeTableS2) and Fig. S\@ref(fig:toeFigureS1)) and geographic region of the neighboring EEZs (Methods; Kruskal-Wallis, $X^2$ = 242.11, DF = 93, *p* < 0.001; Table S3 and Fig. 2A). Overall, most tropical EEZs and stocks will see earlier shifts, with the EEZs of Latin America, the Caribbean and Polynesia experiencing shifts significantly earlier (*p* < 0.05; Tables S\@ref(tab:toeTableS2 and S\@ref(tab:toeTableS3) than almost any other region (Figs. \@ref(fig:toeFigure2)B, S\@ref(fig:toeFigureS1) and S\@ref(fig:toeFigureS2)). In contrast, EEZs and stocks located in temperate regions like northern Europe and eastern Asia are projected to experience much later range shifts. Stocks in countries such as Brazil and New Zealand are not projected to experience range shifts beyond natural variability (Fig. \@ref(fig:toeFigure2)A). On average, shifting transboundary stocks represented 23% (s.d. $\pm$  22%) of yearly fishing revenue from fisheries targeting such stocks within global EEZs between 2004 and 2010 (Fig. \@ref(fig:toeFigure2)A). However, large variation exists, with shifting stocks representing less than 1% of fishing revenue in some countries (e.g., Ireland) and over 90% in others (e.g., Marshall Islands). Moreover, in some EEZs, while few transboundary stocks are projected to shift beyond historical variability, they still represent a large proportion of revenue derived from fisheries within that EEZ (e.g., Peru).

```{r toeFigure2, eval = T,  echo = F, fig.width  = 13,fig.pos= "H", fig.height =11, fig.cap =  "Year of change in the shared distribution of 4,119 transboundary stocks. A) Land polygons show the contribution of shifting stocks to a country or territory’s total fishing revenue from transboundary stocks. Exclusive Economic Zone polygons display average year of range shifts within them. EEZs with no distributional shift between 2006 and 2100 are represented in aqua color. B) Year of shared distribution shifts by UN sub-regions. Points color coded by the median year of range shift. Horizontal dashed line represents the year 2030 – the Sustainable Development Agenda deadline. N = North, S = South, W = West and E = East. Ltn. Ame. and the Car. = Latin America and the Caribbean. Aus and New Z. = Australia and New Zealand."}

@@

We estimated the current (2005 - 2010) contribution of shifting stocks to yearly fishing revenue generated from fisheries targeting transboundary stocks within global EEZs. Within five years of the UN 2030 Agenda, 979 transboundary stocks responsible for 5.3 billion USD in fishing revenue are projected to shift in the EEZs of 76 countries (Fig. \@ref(fig:toeFigure3)); with 4.2 billion USD accounted by 321 stocks shifting in just eleven countries (Fig. \@ref(fig:toeFigure3)).  In terms of the number of transboundary stocks shifting from historical variability, Spain (n = 122), France (n = 85), China and Portugal (both n = 73), and Senegal (n = 66), emerge as the top five countries (Fig. \@ref(fig:toeFigure3)).

```{r toeFigure3, eval = T,  echo = F, fig.width=10, fig.height=8,fig.pos= "H", fig.cap =  "Average range shift of transboundary stocks and corresponding fishing revenue per country. Countries are coded according to different colored symbols representative of 5 distinct regions. Horizontal line separates countries above and below the 75th percentile of annual fishing revenue. Showing country names for those within the top 75th percentile projected to have an average shift within 5 years after the 2030 deadline of the Sustainable Development Agenda.}

@@

@@

<!-- We based the transboundary index on a stock’s centroid defined as the average grid cells that contain the majority of a stock’s abundance (see Methods - Calculating an index of transboundary range shift). Results from the sensitivity analysis suggest that our method is robust to the abundance level (e.g., the top 95%) of the stock (Fig. \@ref(fig:toeFigureS3)). In addition, we set an arbitrary threshold of 2 s.d. (95% confidence) to determine the time of emergence of the transboundary index. Sensitivity tests to a lower threshold of 1 s.d. (for a 68% confidence) resulted in a global average shift in the shared distribution of transboundary stocks by 2029 $\pm$ 27, a difference of about 7 years (Fig. \@ref(fig:toeFigureS4)). In total, 5,745 transboundary stocks would shift their shared distributions between 2006 and 2100 considering a less conservative metric, that is, 33% (n = 1626) more shifting stocks. SENT TO METHODS!!!! -->

We estimated the intensity of the distributional shift of transboundary stocks in terms of changes in catch proportion by 2030 (2020-2040) and 2050 (2040-2060), relative to the recent past (1951-2005). Findings show that by 2030, the global average catch proportion of transboundary stocks that are projected to change beyond the EEZ’s historic proportion is 59% ± 17% (Fig. \@ref(fig:toeFigure4)A). Moreover,  85% (n = 239) of the world EEZs would have experienced changes in catch proportion of transboundary stocks by 2030. Changes in catch proportion are expected to become more common and intense slightly towards the mid -21st century at both the stock and EEZ level (Figs. S\@ref(fig:toeFigureS3) and S\@ref(fig:toeFigureS4)).

```{r toeFigure4, eval = T,  echo = F, fig.width=12, fig.height=10,fig.pos= "H", fig.cap =  "Changes in stock proportion of neighboring Exclusive Economic Zones by 2030 (2021-2040) relative to 1951-2005. Lines represent the average change in transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those gaining shares (arrowhead). Land polygons represent the percentage of stocks that are projected to change their stock share ratio beyond the identified threat point with higher gains identified in warmer colors. Panel B highlights changes for the Caribbean and Northwestern Europe."}

@@

<!-- ![Figure 1. Changes in stock share ratio below each country’s threat point by 2030 (2021-2040) relative to 1951-2005. Lines represent the average change in transboundary stock share ratio with arrows going from EEZ with decreasing stock share (point) to those gaining shares (arrowhead). Land polygons represent the percentage of stocks that are projected to change their stock share ratio beyond the identified threat point with higher gains identified in warmer colors (green to yellow). Panel B zooms into specific areas shown within the gray dashed line boundaries in A.](../Figures/Fig1.png) -->

The direction and intensity of shifts in transboundary stocks are largely related to regional changes in biogeography and the geometry of the EEZs (Fig. \@ref(fig:toeFigure4)A). In some cases, like the Atlantic and Pacific coasts of Northern and Southern America and the Atlantic coast of Southern Africa, the proportion of the  stock is expected to follow a projected poleward shift. However, shifts along the coasts of Pacific Central America and West Africa occur along an equatorial direction. Here, transboundary species follow local oceanographic processes like upwelling zones (e.g., Humboldt current) and patterns of oxygen availability, in addition to temperature gradients. Regions where EEZs are relatively small and have multiple borders (e.g., Caribbean or Northwestern Europe) are expected to have a particularly complex exchange of stock share ratio with no established pattern (Fig. \@ref(fig:toeFigure4)B).

@@

Our findings highlighting early shifts in the shared distribution of transboundary stocks concur with previous studies that have detected and attributed changes in marine catch composition [e.g., @Cheung:2013gk; @Frainer:2017ji; @Last:2011fd]. For example, in the early 2000s, Humboldt squid (*Dosidicus gigas*) substantially expanded its geographic range poleward, reaching the coast of Washington state (US) as a response to climatic, oceanographic and ecological changes [@Zeidberg:2007gs]. A new fishery targeting Humboldt squid quickly developed on the heels of the species’ range expansion [@Pinsky:2014by]. In the northeast Atlantic, fisheries on Atlantic mackerel (*Scomber scombrus*) are multi-laterally managed by the EU, Norway, Iceland, Russia, and Denmark (on behalf of the Faroe Islands and Greenland) through the North-East Atlantic Fisheries Commission (NEAFC). However, a range expansion of Atlantic mackerel into Icelandic waters in 2007 resulted in Iceland capturing 6% of the fishery’s total allowable catch and a further 18% in 2008, without consultation with NEAFC, threatening the sustainability of the stock [@Spijkers:2017ij]. These changes resulted in disputes between Iceland and the Faroe Islands, as well as among NEAFC member states [@Spijkers:2017ij]. As transboundary stocks continue to shift, such international disputes are likely to increase in frequency and requiring a revision of management plans [@Pinsky:2018cb; @Engler:2020bg].

@@

Our results identify regional “hotspots” of climate risk for transboundary fisheries management that will require the prompt adaptation of management plans (Figures \@ref(fig:toeFigure2)A and \@ref(fig:toeFigure4)A). Such regions, like the Caribbean, are characterized by high levels of warming relative to natural internal variability [@IPCC:2019tn; @Hawkins:2012hc] and species’ high vulnerability to warming waters [@IPCC:2019tn]. Moreover, they represent relatively small EEZs that border multiple countries. Game theory predicts that the greater the number of negotiating parties, the harder it is for parties to reach an agreement [@Gronbaek:2020ei], thus, making it particularly challenging for countries in these regions to coordinate the management of shifting transboundary stocks.

Management plans that are not designed to respond to range and abundance shifts will be less resilient to climate change [@Miller:2013iv; @Sumaila:2020tp]. Thus, focusing on these “hotspots” is key to anticipate a potential increase in fisheries conflicts in coming years. Identified strategies to cope with changes in the shared proportion of transboundary stocks include strengthening of current cooperative mechanisms and the consideration of side payments, which do not have to be monetary [@Tunca:2019if; @Miller:2013iv], increased international cooperation [@Miller:2013iv] and management rules that capture distributional shifts [@PalaciosAbrantes:2020vt; @Pinsky:2018cb]. For example, countries that are Party to the Nauru Agreement (PNA) manage their fisheries under a so-called Vessel Day Scheme, which accounts for shifts in stocks and catches, and is therefore somewhat anticipatory [@Aqorau:2018bh]. However, even strategies that take into consideration species’ shifts within the PNA area will likely also have to deal with the ‘newcomer issue’ that NEAFC is currently facing regarding Atlantic mackerel [@Spijkers:2017ij; @Pinsky:2018cb]. Quota allocation methods based on a stock’s current distribution and/or on fixed-historical proportion need to evolve to be more agile [@Sumaila:2020tp; @PalaciosAbrantes:2020vt]. Fisheries based on historic participation could move towards a dynamic method or a combination of both [@PalaciosAbrantes:2020vt; @Sumaila:2020tp]. However, the transition from historic to dynamic allocations can find strong resistance from stakeholders “losing” benefits from a fishery.

@@

The early projected range shifts of a large number of transboundary stocks can be partially attributed to the parallel global emergence of several ocean variables from historical variability e.g., SST is projected to increase beyond historical variability by 2030 in 70% of the global ocean under RCP 8.5 [@Rodgers:2015jq; @Frolicher:2016dg; @Mahlstein:2011fy]. Results from multiple marine ecosystem models show that sea surface temperature (SST) and primary production (NPP) are the main drivers of changes in species distributions across oceanic basins [@BryndumBuchholz:2019kh; @Lotze:2019ce].Specifically, in the species distribution model used here (see methods), SST is the main environmental driver of biomass changes in polar, tropical and upwelling ecosystems while NPP drives temperate regions’ results. The combination of the SST emergence pattern and model characteristics could be partially responsible for the early distributional shift of transboundary stocks in the tropics where marine species live close to their thermal tolerance, making them highly vulnerable to warming waters [@IPCC:2019tn], while also explaining the later, and sometimes non-existent, shifts at higher latitudes (Fig. \@ref(fig:toeFigure2)B).
